---
title: "Notebook 10: Reproduce Scatter Plots"
output: pdf_document
---

# Cap Index vs P-body Enrichment Scatter Plots

Recreate the three reference scatter plots from Jason's analysis using our
independently processed data.

**All plots share:**
- X-axis: `Pub_Log2FC` (published P-body enrichment log2 fold-change)
- Y-axis: `Cap_index` on pseudo-log10 scale
- Magenta dashed horizontal line at Cap_index = 0.1
- Black dashed vertical line at Pub_Log2FC = 0
- GAPDH labeled with black diamond

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(readr)
library(scales)

# Load our processed data (from notebook 08)
dat <- read_csv("../data/processed/08_hub_cage_depmap_impc.csv")
cat(sprintf("Loaded %d genes\n", nrow(dat)))

# Identify the Pub_Log2FC column
pub_col <- grep("Pub_Log2FC|RNA enrichment.*log2", names(dat), value = TRUE)[1]
if (is.na(pub_col)) {
  # Try exact match
  pub_col <- "Pub_Log2FC"
}
cat(sprintf("Using Pub_Log2FC column: %s\n", pub_col))

# Identify the gene name column
gene_col <- grep("Associated Gene Name|gene_name", names(dat), value = TRUE)[1]
cat(sprintf("Using gene name column: %s\n", gene_col))

# Rename for convenience
dat$Pub_Log2FC <- dat[[pub_col]]
dat$gene_name <- dat[[gene_col]]

# Find GAPDH
gapdh <- dat %>% filter(gene_name == "GAPDH")
cat(sprintf("GAPDH: Pub_Log2FC=%.2f, Cap_index=%.4f\n",
            gapdh$Pub_Log2FC[1], gapdh$Cap_index[1]))

# Create output directory
dir.create("../figures", showWarnings = FALSE)
```

## Base plot function

```{r base-plot-fn}
make_cap_scatter <- function(data, highlight_data = NULL,
                              highlight_color = "green",
                              highlight_label = "",
                              cytosol_thresh = 30,
                              title_suffix = "") {
  # Filter to cytosol threshold
  plot_data <- data %>% filter(Cytosol_TPM >= cytosol_thresh)

  # Add small constant for log scale (handle zeros)
  plot_data$Cap_index_plot <- plot_data$Cap_index + 1e-4

  p <- ggplot(plot_data, aes(x = Pub_Log2FC, y = Cap_index_plot)) +
    # Gray background points
    geom_point(color = "gray70", size = 0.5, alpha = 0.5) +

    # Reference lines
    geom_hline(yintercept = 0.1, linetype = "dashed", color = "magenta", linewidth = 0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.3) +

    # Pseudo-log scale for y-axis
    scale_y_continuous(
      trans = pseudo_log_trans(sigma = 0.01, base = 10),
      breaks = c(0, 0.001, 0.01, 0.1, 0.5, 1, 2, 5, 10, 50),
      labels = c("0", "0.001", "0.01", "0.1", "0.5", "1", "2", "5", "10", "50")
    ) +

    labs(
      x = "Published P-body Enrichment (Log2FC)",
      y = "Cap Index (CAGE TPM / Cytosol TPM)",
      title = paste0("Cap Index vs P-body Enrichment", title_suffix),
      subtitle = paste0("Cytosol_TPM >= ", cytosol_thresh)
    ) +

    theme_minimal(base_size = 10) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(size = 11, face = "bold")
    )

  # Add highlighted points
  if (!is.null(highlight_data) && nrow(highlight_data) > 0) {
    highlight_data$Cap_index_plot <- highlight_data$Cap_index + 1e-4
    p <- p + geom_point(
      data = highlight_data,
      aes(x = Pub_Log2FC, y = Cap_index_plot),
      color = highlight_color, size = 0.8, alpha = 0.7
    )
  }

  # Add GAPDH label
  gapdh_plot <- plot_data %>% filter(gene_name == "GAPDH")
  if (nrow(gapdh_plot) > 0) {
    p <- p + geom_point(
      data = gapdh_plot,
      aes(x = Pub_Log2FC, y = Cap_index_plot),
      shape = 18, size = 3, color = "black"
    ) +
    annotate(
      "text",
      x = gapdh_plot$Pub_Log2FC[1] + 0.3,
      y = gapdh_plot$Cap_index_plot[1],
      label = "GAPDH", size = 3, hjust = 0
    )
  }

  return(p)
}
```

## Plot 1: Common Essential Genes (Green)

Filter: Cytosol_TPM >= 30. Common Essential genes highlighted in green.

```{r plot1, fig.width=8, fig.height=6}
essential <- dat %>% filter(DepMap_Common_Essential == 1, Cytosol_TPM >= 30)

p1 <- make_cap_scatter(
  dat,
  highlight_data = essential,
  highlight_color = "darkgreen",
  highlight_label = "Common Essential",
  cytosol_thresh = 30,
  title_suffix = " — Common Essential (green)"
)

print(p1)
ggsave("../figures/CapIndex_v_PubLog2FC_CommonEssential.pdf", p1,
       width = 8, height = 6)
```

## Plot 2: Strongly Selective Genes (Orchid)

Filter: Cytosol_TPM >= 30. Strongly Selective genes highlighted in orchid.

```{r plot2, fig.width=8, fig.height=6}
selective <- dat %>% filter(DepMap_Strongly_Selective == 1, Cytosol_TPM >= 30)

p2 <- make_cap_scatter(
  dat,
  highlight_data = selective,
  highlight_color = "orchid",
  highlight_label = "Strongly Selective",
  cytosol_thresh = 30,
  title_suffix = " — Strongly Selective (orchid)"
)

print(p2)
ggsave("../figures/CapIndex_v_PubLog2FC_StronglySelective.pdf", p2,
       width = 8, height = 6)
```

## Plot 3: Embryonic Lethal Orthologs (Red)

Filter: Cytosol_TPM >= 5. IMPC embryonic lethal orthologs highlighted in red.

```{r plot3, fig.width=8, fig.height=6}
lethal <- dat %>% filter(IMPC_Embryonic_Lethal_Ortholog == 1, Cytosol_TPM >= 5)

p3 <- make_cap_scatter(
  dat,
  highlight_data = lethal,
  highlight_color = "red",
  highlight_label = "Embryonic Lethal",
  cytosol_thresh = 5,
  title_suffix = " — Embryonic Lethal Orthologs (red)"
)

print(p3)
ggsave("../figures/CapIndex_v_PubLog2FC_EmbryonicLethal.pdf", p3,
       width = 8, height = 6)
```

## Summary

```{r summary}
cat("Plots saved to ../figures/\n")
cat("Compare against reference PDFs in ../reference/\n")
cat(sprintf("\nPlot 1: %d common essential genes highlighted (Cytosol >= 30)\n",
            nrow(essential)))
cat(sprintf("Plot 2: %d strongly selective genes highlighted (Cytosol >= 30)\n",
            nrow(selective)))
cat(sprintf("Plot 3: %d embryonic lethal orthologs highlighted (Cytosol >= 5)\n",
            nrow(lethal)))
```
